---
title: Plotmath - how to get scientific notation on plots
author: Fiona Seaton
date: 2018-11-20

draft: false
type: docs

linktitle: Plotmath
menu:
  tutorial:
    parent: Plotting
    weight: 15
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



This is a quick section on how to get weird symbols onto your plot labels, mainly because I want to write the rules down somewhere before I forget. R has built in functionality to put all sorts of weird mathematical expressions on plots, see the help page for plotmath for a long list of things that are possible. I'm using expression to create the labels because it seems to be the simplest option I've found so far.

```{r regex table, echo = FALSE, message = FALSE, warnings=FALSE, results='asis'}
reg_table <- data.frame(Character = c("~","\\*","^","[]"),
                        Meaning = c("separate with a space","put next to each other with no space","put the next bit in superscript",
                                    "put the bit in brackets in subscript"))
knitr::kable(reg_table)
```



```{r plot super and sub}
plot(cars, xlab = expression("Speed (hr"^-1*")"), ylab = expression(bold(Distance[30])~"(ft)"))
```

```{r plotmath 2}
par(mar = c(5,5,4,2) + 0.1)
plot(cars, xlab = expression(pi~and~tilde(20)*degree), ylab = expression(infinity%down%N[2]*O-N))
```

You can also reference unicode symbols directly, but not all are supported. I'm using "\u2030" here to show the per mille symbol.
```{r plotmath 3}
par(mar = c(5,5,4,2) + 0.1)
plot(cars, 
     ylab = expression(Delta^13*C-CO[2]~"(\u2030)"),  
     xlab = expression(italic(A[380])~"("*mu*"mol"~CO[2]~m^-2~s^-1*")"))
```

All of the above are possible in ggplot2, as shown below. You can also plot symbols directly onto the plot using expression within the labels value in geom_text.
```{r ggplot}
library(ggplot2)
dat <- data.frame(X = gl(7,1),
                  Y = runif(7))

ggplot(dat, aes(x = X, y =Y)) + geom_bar(stat = "identity") +
  geom_text(aes(y = Y + 0.1),
            label = c(expression(alpha), 
                      expression(beta), expression(gamma), 
                      expression(delta), expression(epsilon), 
                      expression(zeta), expression(eta))) +
  labs(x = expression(infinity %~~% frac(Upsilon1, widetilde(alpha*beta))),
       y = expression(Omega*bolditalic("(\u2020)")))
```


### emojis because why not
Also here is how to get emoji onto plots using the emojifont package, because I find this far too amusing. These examples are lifted directly from [this comment on stack overflow](https://stackoverflow.com/a/47763512) and Guangchuang Yu's [vignette on phylomoji](https://guangchuangyu.github.io/software/ggtree/vignettes/phylomoji.html).

```{r emojifont}
library(emojifont)

# data setup
df <- data.frame(name_emoji = c("closed book", "confused face","dizzy", 
                                "face with tears of joy", "flexed biceps", "light bulb"), 
                      n = c(1L, 7L, 20L, 1L, 1L, 1L), 
                      emoji = c("\U0001f4d5", "\U0001f615", "\U0001f4ab", 
                                "\U0001f602", "\U0001f4aa", "\U0001f4a1"))

ggplot(df, aes(name_emoji, n, label = emoji)) + 
  geom_bar(stat = "identity") +
  geom_text(family = "EmojiOne", size = 6, vjust = -.5) +
  scale_y_continuous(limits = c(0,25)) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

```{r phylomoji}
library(ggtree)
tree_text <- "(((((cow, (whale, dolphin)), (pig2, boar)), camel), fish), seedling);"
x <- read.tree(text=tree_text)
ggtree(x, linetype="dashed", color='firebrick') +
  xlim(NA, 7) + ylim(NA, 8.5) +
  geom_tiplab(aes(color=label), parse='emoji', size=14, vjust=0.25) +
  labs(caption="powered by ggtree + emojifont")
```

