---
title: Summary tables and statistics
author: Fiona Seaton
date: 2018-10-09

draft: false
type: docs

output:
  blogdown::html_page:
    toc: true

linktitle: Summary statistics
menu:
  tutorial:
    parent: R club
    weight: 2
---

This is a continuation of the tutorial on data manipulation, with many examples of different summary tables and statistics. It's largely using the janitor, summarytools and psych packages which have loads of great functions for quickly examining your data and I highly recommend them!


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


First of all we need to load up the packages we want to use today

```{r packages}
library(readr)   ## for reading in data
library(dplyr)   ## for data manipulation
library(janitor) ## for data cleaning tools
library(psych)   ## for some useful data exploration graphs
library(summarytools) ## for some summary statistics and tables
```

We're using the same data as the data manipulation tutorial to show a variety of different summary statistics and tables
```{r data}
CS <- read_csv("CS_fake.csv")
CS <- clean_names(CS)
CS
```



## Summary tables and statistics
There's a _lot_ of different functions in R for summarising dataframes, so we're going to go through a few useful ones

The summarytools package has a few useful functions, such as the summary table produced below:
```{r summarytools, results = 'asis'}
print(dfSummary(CS[,2:6], style = 'grid', graph.magnif = 0.85), method = "render", omit.headings = TRUE)
```


### Frequency tables and cross-tabulating factors
Both janitor and summarytools offer functions for frequency tables and cross-tabulating factors. 

```{r frequency tables summarytools, results = 'asis'}
freq(CS$habitat, style = "rmarkdown", omit.headings = TRUE) ##summarytools
print(ctable(CS$country, CS$county, prop = "n"), method = "render", omit.headings = TRUE) ##summarytools
```


For me janitor is a little more intuitive, and is also built to be compatible with the pipe operator

```{r frequency tables}
tabyl(CS, habitat)
tabyl(CS, country, county) 
tabyl(CS, country, habitat, county)
```

### Summary statistics by groups
dataframes can be described by a categorical variable within both summarytools and psych.
The describeBy function in psych is simple to write but only likes numeric variables. I also don't know of any way to pick and choose which stats are reported.

```{r dataframe grouping variables}
describeBy(CS, "country")
```

summarytools' version is longer to write but appears to be more flexible. I've set some parameters here to make it print better on this document - you may want to remove the style and method arguments to make it look nicer when working from the console.

```{r grouping summarytools, results = 'asis'}
## summarytools's version is longer to write but more flexible
# First save the results
CS_stats_by_country <- by(data = CS, 
                          INDICES = CS$country, 
                          FUN = descr, stats = c("mean", "sd", "min", "med", "max"), 
                          transpose = TRUE, style = "rmarkdown", omit.headings = TRUE)
# then view
view(CS_stats_by_country, method = "pander")
```

You can do the above using group_by() then summarise() from the dplyr package but as far as I know you'll have to manually write out which stats you want which can be a hassle:

```{r dataframe grouping summary dplyr}
CS %>% group_by(country) %>%
  summarise(mean_pH = mean(ph, na.rm = TRUE),
            mean_carbon = mean(carbon_percent, na.rm = TRUE))
```




## Quick graphs
While summary statistic tables can be very useful they are no substitute for plotting out your data. For an interesting discussion about this (with dinosaurs!) see [this page on how different datasets can be with the same summary statistics.](https://www.autodeskresearch.com/publications/samestats)

Histograms are great for seeing what kind of distributions your variables have, I often use the multi.hist function from within the psych package to look at my data quickly and easily. Note that this function doesn't always return the plotting environment to what it was previously so we specify that the plotting parameters should return to one plot per page afterwards

```{r histogram}
multi.hist(CS[,4:6])
par(mfrow=c(1,1)) 
```

The pairs plot within R, and the extensions upon it from many other package, is a great way to see how the variables from your dataset are related to each other.The pairs.panels function in psych shows the name of each variable and its histogram down the diagonal, the scatterplots on the lower triangle and the correlation coefficient on the upper triangle. You can change the type of correlation coefficient, and various plotting parameters easily.

```{r pairsplot}
pairs.panels(CS[,4:6])
```

If you want to take an in depth look at one of the plots and identify certain points then you can plot the scatterplot and use the identify function to interactively select points and get a label for what they are - this doesn't work in rmarkdown but you can try it yourself.

```{r identify, results = "hide"}
plot(nitrogen_percent ~ carbon_percent, CS)
identify(CS$carbon_percent, CS$nitrogen_percent, labels=CS$plot_id)
```

Other useful plots include coplots in base R, which show how the relationship between two variables changes with a third variable
```{r coplot, results = "hide"}
coplot(ph ~ carbon_percent|nitrogen_percent, CS)
```


That's all that comes to mind right now. There are __*so many*__ different ways of manipulating and exploring datasets in R. These are just a few that I have found useful and I hope you do too!
